---
title: "Layers overview"
format: html
editor: visual
---

### Using layers: an overview

In Mapbox GL JS and MapLibre, datasets are added to maps as *sources* which are then styled as *layers*. **mapgl** aims to expose the sources and layers APIs to R users in ways that honor the deep customization available in the JavaScript libraries but also accommodate R users' typical workflows.

Geospatial practitioners in R will typically work with objects from the **sf** package. The initial release of **mapgl** natively supports `sf` objects, and will aim to support other geospatial formats (such as objects from the **terra** package) in the future.

Objects of class `sf` can be specified as sources for the map either through the `add_source()` function or via the `source` parameter in one of **mapgl**'s layer functions. The `add_fill_layer()` function calls the Mapbox GL JS `addLayer()` function internally with a `fill` type, and enumerates the available options for styling the layer as function arguments.

**mapgl** users will often want to use the `fit_bounds(animate = FALSE)` function to fix the map view to a given layer's bounding box.

```{r, message = FALSE}

library(mapgl)
library(sf)

nc <- st_read(system.file("shape/nc.shp", package="sf"))

mapboxgl() |> 
  fit_bounds(nc, animate = FALSE) |> 
  add_fill_layer(id = "nc_data",
                 source = nc,
                 fill_color = "blue",
                 fill_opacity = 0.5)
```

An overview of available layers in **mapgl** are below. Layers can be used with either `mapboxgl()` or `maplibre()` maps.

### Line layers

```{r}
library(mapgl)
library(tigris)
options(tigris_use_cache = TRUE)

loving_roads <- roads("TX", "Loving")

maplibre(style = maptiler_style("backdrop")) |>
  fit_bounds(loving_roads, animate = FALSE) |>
  add_line_layer(
    id = "roads",
    source = loving_roads,
    line_color = "navy",
    line_opacity = 0.7
  )
```

### Circle layers

At the time of this writing, circle clustering is not yet implemented; however, this is on the roadmap for the package.

```{r, message = FALSE}
library(mapgl)
library(sf)
library(dplyr)

# Set seed for reproducibility
set.seed(1234)

# Define the bounding box for Washington DC (approximately)
bbox <- st_bbox(c(
  xmin = -77.119759,
  ymin = 38.791645,
  xmax = -76.909393,
  ymax = 38.995548
),
crs = st_crs(4326))

# Generate 30 random points within the bounding box
random_points <- st_as_sf(
  data.frame(
    id = 1:30,
    lon = runif(30, bbox["xmin"], bbox["xmax"]),
    lat = runif(30, bbox["ymin"], bbox["ymax"])
  ),
  coords = c("lon", "lat"),
  crs = 4326
)

# Assign random categories
categories <- c('music', 'bar', 'theatre', 'bicycle')
random_points <- random_points %>%
  mutate(category = sample(categories, n(), replace = TRUE))

# Map with circle layer
mapboxgl(style = mapbox_style("dark")) %>%
  fit_bounds(random_points, animate = FALSE) %>%
  add_circle_layer(
    id = "poi-layer",
    source = random_points,
    circle_color = match_expr(
      "category",
      values = c("music", "bar", "theatre",
                 "bicycle"),
      stops = c("#1f78b4", "#33a02c",
                "#e31a1c", "#ff7f00")
    ),
    circle_radius = 8,
    circle_stroke_color = "#ffffff",
    circle_stroke_width = 2,
    circle_opacity = 0.8,
    tooltip = "category",
    hover_options = list(circle_radius = 12,
                         circle_color = "#ffff99")
  ) %>%
  add_categorical_legend(
    legend_title = "Points of Interest",
    values = c("Music", "Bar", "Theatre", "Bicycle"),
    colors = c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00"),
    circular_patches = TRUE
  )
```

### Symbol layers

Symbol layers offer a wide range of arguments for customizing icon and label appearance; however not all arguments will work with all icons. The `icon_image` argument will look for a string that represents an icon found in the map style's sprite. [Read more about sprites here](https://docs.mapbox.com/style-spec/reference/sprite/).

```{r}
mapboxgl(style = mapbox_style("light")) |>
  fit_bounds(random_points, animate = FALSE) |>
  add_symbol_layer(
    id = "points-of-interest",
    source = random_points,
    icon_image = get_column("category"),
    icon_allow_overlap = TRUE,
    tooltip = "category"
  )
```

### Heatmap layers

Heatmap layers take an object of geometry type POINT and visualize the density of those points in a visually attractive way. `add_heatmap_layer()` takes **sf** POINT objects; the example below shows how to read in a remote GeoJSON file as a source as well.

```{r}
library(mapgl)

mapboxgl(style = mapbox_style("dark"),
         center = c(-120, 50),
         zoom = 2) |>
  add_heatmap_layer(
    id = "earthquakes-heat",
    source = list(
      type = "geojson",
      data = "https://docs.mapbox.com/mapbox-gl-js/assets/earthquakes.geojson"
    ),
    heatmap_weight = interpolate(
      column = "mag",
      values = c(0, 6),
      stops = c(0, 1)
    ),
    heatmap_intensity = interpolate(
      property = "zoom",
      values = c(0, 9),
      stops = c(1, 3)
    ),
    heatmap_color = interpolate(
      property = "heatmap-density",
      values = seq(0, 1, 0.2),
      stops = c('rgba(33,102,172,0)', 'rgb(103,169,207)',
                'rgb(209,229,240)', 'rgb(253,219,199)',
                'rgb(239,138,98)', 'rgb(178,24,43)')
    ),
    heatmap_opacity = 0.7
  )
```

### Fill-extrusion layers

```{r}
library(mapgl)

maplibre(
  style = maptiler_style("basic"),
  center = c(-74.0066, 40.7135),
  zoom = 15.5,
  pitch = 45,
  bearing = -17.6
) |>
  add_vector_source(
    id = "openmaptiles",
    url = paste0("https://api.maptiler.com/tiles/v3/tiles.json?key=",
                 Sys.getenv("MAPTILER_API_KEY"))
  ) |>
  add_fill_extrusion_layer(
    id = "3d-buildings",
    source = 'openmaptiles',
    source_layer = 'building',
    fill_extrusion_color = interpolate(
      column = 'render_height',
      values = c(0, 200, 400),
      stops = c('lightgray', 'royalblue', 'lightblue')
    ),
    fill_extrusion_height = list(
      'interpolate',
      list('linear'),
      list('zoom'),
      15,
      0,
      16,
      list('get', 'render_height')
    )
  )
```

### Raster layers

**mapgl** does not natively support R rasters in its initial release, but aims to do so in the future. Raster sources can be added through `add_image_source()` for remotely-hosted image files, or `add_raster_source()` for remotely-hosted raster tiles.

```{r}
mapboxgl(style = mapbox_style("dark"),
         zoom = 5,
         center = c(-75.789, 41.874)) |>
  add_image_source(
    id = "radar",
    url = "https://docs.mapbox.com/mapbox-gl-js/assets/radar.gif",
    coordinates = list(
      c(-80.425, 46.437),
      c(-71.516, 46.437),
      c(-71.516, 37.936),
      c(-80.425, 37.936)
    )
  ) |>
  add_raster_layer(
    id = 'radar-layer',
    source = 'radar',
    raster_fade_duration = 0
  )
```

### Markers

Markers represent a unique visual component in Mapbox GL JS and MapLibre GL JS, as they highlight locations but do not count as map *layers*. In **mapgl**, users can add markers using the `add_markers()` function. A single marker can be added as a length-2 vector of longitude and latitude; a list of length-2 vectors or an **sf** POINT object will add multiple markers.

```{r}
mapboxgl(
  style = mapbox_style("streets"),
  center = c(-74.006, 40.7128),
  zoom = 10
) |> 
add_markers(
  c(-74.006, 40.7128),
  color = "blue",
  rotation = 45,
  popup = "A marker"
)
```
