---
title: "Fundamentals of map design with mapgl"
format:
  html:
    standalone: false
editor: visual
---

```{r setup, echo = FALSE}
options(tigris_use_cache = TRUE)
knitr::opts_chunk$set(fig.height = 4)

```

The **mapgl** package aims to expose the powerful map design capabilities of Mapbox GL JS and Maplibre GL JS, while still feeling intuitive to R users. This means that map-making may require a little more code than other mapping packages - but it also gives you maximum flexibility in how you design your maps.

Let's grab some data from **tidycensus** on median age by Census tract in Florida and initialize an empty map focused on Florida.

```{r}
library(tidycensus)
library(mapgl)

fl_age <- get_acs(
  geography = "tract",
  variables = "B01002_001",
  state = "FL",
  year = 2024,
  geometry = TRUE
)

fl_map <- mapboxgl(mapbox_style("light"),
                   bounds = fl_age)

fl_map
```

### Continuous styling

Styling in Mapbox GL JS and Maplibre GL JS is typically handled through *expressions*. Expressions allow for quite a bit of customization for map-makers, but can feel clunky for R users. **mapgl** includes several functions to help R users translate their code into expressions for use in their data visualizations.

For continuous color scales, the `interpolate_palette()` function automatically calculates appropriate break points and creates smooth color transitions. You can specify the classification method ("equal", "quantile", or "jenks") and either a palette function or specific colors:

```{r}
# Automatic continuous scale with equal breaks
continuous_scale <- interpolate_palette(
  data = fl_age,
  column = "estimate",
  method = "equal",
  n = 5,
  palette = viridisLite::plasma
)

fl_map |>
  add_fill_layer(
    id = "fl_tracts",
    source = fl_age,
    fill_color = continuous_scale$expression,
    fill_opacity = 0.5
  ) |>
  add_legend(
    "Median age in Florida",
    values = get_legend_labels(continuous_scale, digits = 0),
    colors = get_legend_colors(continuous_scale),
    type = "continuous"
  )
```

For lower-level control, the `interpolate()` function creates an `interpolate` expression where you manually specify values and stops. This gives you complete control over the color mapping:

```{r}
fl_map |>
  add_fill_layer(
  id = "fl_tracts",
  source = fl_age,
  fill_color = interpolate(
    column = "estimate",
    values = c(20, 80),
    stops = c("lightblue", "darkblue"),
    na_color = "lightgrey"
  ),
  fill_opacity = 0.5
 ) |>
  add_legend(
    "Median age in Florida",
    values = c(20, 80),
    colors = c("lightblue", "darkblue")
  )
```

You can also let readers switch among several continuous color ramps directly from the legend. The map layer is styled with the selected ramp, and clicking a different ramp in the legend updates both the legend gradient and the map.

```{r}
age_ramps <- list(
  Viridis = viridisLite::viridis(5),
  Magma = viridisLite::magma(5),
  Blue = c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c"),
  Fire = c("#ffffcc", "#ffeda0", "#feb24c", "#f03b20", "#bd0026")
)

age_picker_scale <- interpolate_palette(
  data = fl_age,
  column = "estimate",
  method = "equal",
  n = 5,
  color_ramps = age_ramps,
  selected_ramp = "Viridis",
  na_color = "lightgrey"
)

maplibre(
  style = carto_style("positron"),
  bounds = fl_age
) |>
  add_fill_layer(
    id = "fl_age_picker",
    source = fl_age,
    fill_color = age_picker_scale$expression,
    fill_opacity = 0.65,
    fill_outline_color = "rgba(255,255,255,0.25)"
  ) |>
  add_legend(
    "Median age in Florida",
    values = get_legend_labels(age_picker_scale, digits = 0),
    colors = age_picker_scale,
    type = "continuous",
    layer_id = "fl_age_picker",
    ramp_picker = TRUE,
    style = legend_style(
      background_color = "white",
      background_opacity = 0.92,
      border_color = "#d1d5db",
      border_width = 1,
      shadow = TRUE
    )
  )
```

### Categorical styling

Cartographers may prefer a binned method for visualizing their data rather than the continuous palette shown above. **mapgl** includes several classification functions that automate common choropleth workflows. The `step_equal_interval()` function creates equal interval breaks, while `step_quantile()` and `step_jenks()` create quantile and Jenks natural breaks respectively.

These functions automatically calculate appropriate break points and generate step expressions. You can specify the number of classes and the color palette to use:

```{r}
# Automatic quantile classification with 5 classes
q_class <- step_quantile(
  data = fl_age,
  column = "estimate",
  n = 5,
  colors = RColorBrewer::brewer.pal(5, "PRGn")
)

fl_map |>
  add_fill_layer(
    id = "fl_tracts",
    source = fl_age,
    fill_color = q_class$expression,
    fill_opacity = 0.5
  ) |>
  add_legend(
    "Median age in Florida",
    values = get_legend_labels(q_class, digits = 0, suffix = " years"),
    colors = get_legend_colors(q_class),
    type = "categorical"
  )
```

For lower-level control, you can still use the `step_expr()` function directly. Step expressions require a `base` value followed by a series of `stops` and threshold `values`:

```{r}
brewer_pal <- RColorBrewer::brewer.pal(5, "RdYlBu")

fl_map |>
  add_fill_layer(
  id = "fl_tracts",
  source = fl_age,
  fill_color = step_expr(
    column = "estimate",
    base = brewer_pal[1],
    stops = brewer_pal[2:5],
    values = seq(25, 70, 15),
    na_color = "white"
  ),
  fill_opacity = 0.5
 ) |>
  add_legend(
    "Median age in Florida",
    values = c(
      "Under 25",
      "25-40",
      "40-55",
      "55-70",
      "Above 70"
    ),
    colors = brewer_pal,
    type = "categorical"
  )
```

### Pop-ups, tooltips, and highlighting

Mapmakers will often want to expose some additional interactivity to their users in the form of on-click popups, hover tooltips, and other hover effects. In native JavaScript, this can be tricky to set up as it requires knowledge of events, queries, and feature states in these libraries. **mapgl** wraps this functionality to make these features accessible to R users.

The `popup` and `tooltip` arguments take a string as input representing the name of the column to display on click or on hover. Both arguments accommodate HTML, so the best way to set this up is to create a column of values to display in the popup or tooltip, then use this column when adding the layer.

Hover effects can be set with the `hover_options` argument. This argument takes a list of key-value pairs where the keys are arguments for a given layer type (in this case, the fill layer) and arguments are the desired values on hover. In the example shown here, we tell Mapbox GL JS to change a Census tract's fill to yellow and fill opacity to 1 when the users hovers over the tract.

```{r}
fl_age$popup <- glue::glue(
  "<strong>GEOID: </strong>{fl_age$GEOID}<br><strong>Median age: </strong>{fl_age$estimate}"
)

fl_map |>
  add_fill_layer(
  id = "fl_tracts",
  source = fl_age,
  fill_color = interpolate(
    column = "estimate",
    values = c(20, 80),
    stops = c("lightblue", "darkblue"),
    na_color = "lightgrey"
  ),
  fill_opacity = 0.5,
  popup = "popup",
  tooltip = "estimate",
  hover_options = list(
    fill_color = "yellow",
    fill_opacity = 1
  )
 ) |>
  add_legend(
    "Median age in Florida",
    values = c(20, 80),
    colors = c("lightblue", "darkblue")
  )
```

### Bivariate styling

Bivariate maps visualize two variables at once by assigning each feature to a joint class. In this example, we'll compare median age and median household income for Florida Census tracts.

The built-in bivariate palettes are 3-by-3 matrices. Rows represent the y-variable from low to high, and columns represent the x-variable from low to high. You can inspect the available palettes with `names(bivariate_palettes())`, or pass your own 3-by-3 matrix to the `colors` argument in `bivariate_scale()`. By default, `bivariate_scale()` uses quantile breaks; pass `x_breaks` and `y_breaks` as four-value numeric vectors when you need stable bins across maps or filtered views.

```{r}
fl_age_income <- get_acs(
  geography = "tract",
  variables = c(
    median_age = "B01002_001",
    median_income = "B19013_001"
  ),
  state = "FL",
  year = 2024,
  geometry = TRUE,
  output = "wide"
)

names(fl_age_income) <- sub("E$", "", names(fl_age_income))

fl_age_income$popup <- glue::glue(
  "<strong>GEOID: </strong>{fl_age_income$GEOID}<br>",
  "<strong>Median age: </strong>{round(fl_age_income$median_age, 1)}<br>",
  "<strong>Median household income: </strong>${format(round(fl_age_income$median_income), big.mark = ',')}"
)

age_income_scale <- bivariate_scale(
  data = fl_age_income,
  x = "median_income",
  y = "median_age", na_color = "white", palette = "blue_red"
)

maplibre(
  style = carto_style("positron"),
  bounds = fl_age_income
) |>
  add_fill_layer(
    id = "fl_age_income",
    source = fl_age_income,
    fill_color = age_income_scale$expression,
    fill_opacity = 0.75,
    fill_outline_color = "rgba(255,255,255,0.25)",
    popup = "popup",
    hover_options = list(
      fill_opacity = 1
    )
  ) |>
  add_bivariate_legend(
    age_income_scale,
    legend_title = "Age and income",
    x_title = "Higher income",
    y_title = "Older median age",
    layer_id = "fl_age_income",
    style = legend_style(
      background_color = "white",
      background_opacity = 0.94,
      border_color = "#d1d5db",
      border_width = 1,
      shadow = TRUE
    )
  )
```
